Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-07-30 pirouette_article.tex
Changes @rsetienne

\begin{figure}[H]
  \includegraphics[width=0.9\textwidth]{pirouette_example_30/errors.png}
  \caption{
    \new{
      The impact of the tree prior for 
      the example tree in Figure \ref{fig:dd_tree}. 
      The alignment for this true tree was generated using a JC substitution model 
      and strict clock model. 
      For inferring the tree from this alignment in the 'generative' scenario,
      the same substitution and clock models were used, 
      and a Yule tree prior (this is the assumed generative model,
      because the real generative model is of course unknown).
      For the twin tree, the same inference model is used.
      In the 'best' scenario, for the true tree, 
      the best-fitting candidate models on the true tree
      (JC substitution model, RLN clock model and BD tree prior) 
      was used. 
      In the 'best' scenario, for the twin tree, 
      the best-fitting candidate models on the twin tree
      (JC substitution model, RLN clock model and Yule tree prior) 
      was used. 
      The twin distributions show the baseline inference error.
    }
    Vertical dashed lines show the median error value per distribution.
  }
  \label{fig:example_30}
\end{figure}

@@

Comparing the upper and lower panel of Figure \ref{fig:example_30}, 
we can see that 
\new{
  the best candidate model is slightly worse at inferring the true tree,
  than the (assumed) generative model, 
  indicating that the generative inference model we selected is a good pick.
}

@@

Interpreting the results of \verb;pirouette; is up to the user; 
\verb;pirouette; does not answer the question 
whether the inference error is too large to trust the inferred tree. The user is encouraged to use different statistics to measure the error. The nLTT statistic is
a promising starting point, as it can compare any two trees and 
results in an error distribution of known range, but one may also explore other statistics.
In principle, \verb;pirouette; allows for this, but in our example we used a diversification model (DD) that only deviates from the Yule and BD models in the temporal branching pattern, not in the topology.
\new{
  For models that make different predictions on topology, 
  the twinning process should be modified.
}
\new{
	As noted in the introduction, Duch\^{e}ne and 
	colleagues [\cite{duchene2018phylodynamic}]
	also developed a method to assess the adequacy of a tree model
	on empirical trees. They simulated trees from the posterior distribution of 
	the parameters and then compared this to the originally inferred tree using 
	tree statistics, to determine whether the assumed tree model in inference 
	indeed generates the tree as inferred. This is useful if these trees match, 
	but when they do not, this does not mean that the inferred tree is incorrect; 
	if sufficient data is available the species tree prior may not be important,
	and hence inference may be adequate 
	even though the assumed species tree prior is not. 
	In short, the approach is applied to empirical trees and 
	compares the posterior and prior distribution of trees (with the latter 
	generated with the posterior parameters!).
    By contrast,}
    \verb;pirouette;
	\new{aims to expose when assuming standard priors for the 
	species tree are a mis- or underparameterization. 
	Hence, our approach applies to simulated trees and compares 
	the posterior distributions of trees generated with 
	a standard and non-standard model, but inferred with a standard one. 
	The two methods therefore complement one another.
}
Furthermore, we note that the \verb;pirouette; pipeline is not restricted 
to exploring the effects of a new species tree model. 
The pipeline can also be used to explore the effects of non-standard 
clock or site models, such as relaxed clock models with a non-standard 
distribution, correlated substitutions on sister lineages, or elevated 
substitutions rates during speciation events. 
It is, however, beyond the scope of this paper to discuss all these options 
in more detail.

In conclusion, \verb;pirouette; can show the errors to be expected
when the model assumed in inference is different from the actual generative model.
The user can then judge whether or not this new model 
should be implemented in the Bayesian phylogenetic tool.

>>>>

\begin{figure}[H]
  \includegraphics[width=0.9\textwidth]{pirouette_example_30/errors.png}
  \caption{
    \new{
      The impact of the tree prior for 
      the example tree in Figure \ref{fig:dd_tree}. 
      The alignment for this true tree was generated using a JC substitution model and strict clock model. 
      For inferring the tree from this alignment in the 'generative' scenario,
      the same substitution and clock models were used, 
      and a Yule tree prior (this is the assumed generative model,
      because the real generative model is assumed to be unknown).
      For the twin tree, the same inference models were used.
      In the 'best' scenario, for the true tree, 
      the best-fitting candidate models were JC substitution model, RLN clock model and BD tree prior, while for the twin tree, the best-fitting candidate models were JC substitution model, RLN clock model and Yule tree prior. 
      The twin distributions show the baseline inference error.
    }
    Vertical dashed lines show the median error value per distribution.
  }
  \label{fig:example_30}
\end{figure}

@@

Comparing the upper and lower panel of Figure \ref{fig:example_30}, 
we can see that 
\new{
  the best candidate model is slightly worse at inferring the true tree,
  than the (assumed) generative model, 
  indicating that the generative inference model we selected is a good choice.
}

@@

Interpreting the results of \verb;pirouette; is up to the user; 
\verb;pirouette; does not answer the question 
whether the inference error is too large to trust the inferred tree. The user is encouraged to use different statistics to measure the error. The nLTT statistic is
a promising starting point, as it can compare any two trees and 
results in an error distribution of known range, but one may also explore other statistics, for example statistics that depend on the topology of the tree,
While \verb;pirouette; allows for this in principle, in our example we used a diversification model (DD) that only deviates from the Yule and BD models in the temporal branching pattern, not in the topology.
\new{
  For models that make different predictions on topology, 
  the twinning process should be modified.
}
\new{
	As noted in the introduction, Duch\^{e}ne and 
	colleagues [\cite{duchene2018phylodynamic}]
	also developed a method to assess the adequacy of a tree model
	on empirical trees. They simulated trees from the posterior distribution of 
	the parameters and then compared this to the originally inferred tree using 
	tree statistics, to determine whether the assumed tree model in inference 
	indeed generates the tree as inferred. This is useful if these trees match, 
	but when they do not, this does not mean that the inferred tree is incorrect; 
	if sufficient data is available the species tree prior may not be important,
	and hence inference may be adequate 
	even though the assumed species tree prior is not. 
	In short, the approach is applied to empirical trees and 
	compares the posterior and prior distribution of trees (with the latter 
	generated with the posterior parameters!).
    By contrast,}
    \verb;pirouette;
	\new{aims to expose when assuming standard priors for the 
	species tree are a mis- or underparameterization. 
	Hence, our approach applies to simulated trees and compares 
	the posterior distributions of trees generated with 
	a standard and non-standard model, but inferred with a standard one. 
	The two methods therefore complement one another.
}
Furthermore, we note that the \verb;pirouette; pipeline is not restricted 
to exploring the effects of a new species tree model. 
The pipeline can also be used to explore the effects of non-standard 
clock or site models, such as relaxed clock models with a non-standard 
distribution, correlated substitutions on sister lineages, or elevated 
substitutions rates during speciation events. 
It is, however, beyond the scope of this paper to discuss all these options 
in more detail.

In conclusion, \verb;pirouette; can show the errors in phylogenetic reconstruction expected
when the model assumed in inference is different from the actual generative model.
The user can then judge whether or not this new model 
should be implemented in a Bayesian phylogenetic tool.

### 2020-07-30 pirouette_supplement.tex
Changes @rsetienne

The ESSes of the 'true' pipeline for the best candidate model
are shown in Table \ref{tab:esses_best}. 
From the names of the estimated parameters, it is clear that
the best candidate model has had a JC nucleotide substitution 
model (no parameter needed to be estimated)
an RLN clock model (which can be inferred from the parameter 'rate.mean') 
and a BD model ('BirthDeath', 'BDBirthRate' and 'BDDeathRate').

>>>>

The ESSes of the 'true' pipeline for the best candidate model
are shown in Table \ref{tab:esses_best}. 
From the names of the estimated parameters, it is clear that
the best candidate model has a JC nucleotide substitution 
model (no parameter needed to be estimated)
an RLN clock model (which can be inferred from the parameter 'rate.mean') 
and a BD tree prior ('BirthDeath', 'BDBirthRate' and 'BDDeathRate').
